---
title: "SFA1 - DOC workup samples 52-102, B2"
author: "Cassi Wattenburger"
date: "3/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(results = "hide")
```

**Goal of script:** Work up DOC data from sand microcosms and identify high and low producing micrcosms for next pahse of experiment.

Script based heavily on Taylor Cyle's script, shimadzu_workup_template.Rmd.

Re-scripting/organizing it myself to make sure I understand everything.

Meant to be run with samples analyzed using Shimadzu NPOC measurement only, 3/5 injections.

### Script requirements
* Shimadzu exported data with header info and blank rows removed (first 26 lines of .txt file)
* Sample metadata file with atleast an "ID"id column and "dilution" column (ie 8 if you diluted samples by hand 8-fold prior to measuring NOT the same as auto-dilute which the machine performs for you)

### Run specific info
* NPOC standard = 100 ppm KHP
* Ext check 1 = 1 mM glycine
* Ext check 2 = 1 mM alanine
* Ext check 3 = 1 mM glucose

## Step 0: Clear workspace, attach packages, set working directory
```{r}
rm(list=ls())

library(plyr)
library(tidyverse)
require(broom)

# Set working directory to location of files
setwd("~/SFAgrowthrate/SFAphase1/DOC")
```

```{r, results="show"}
sessionInfo()
```

```{r, echo=FALSE}
# Names of external/internal checks and samples as entered into Shimadzu software
stds.name <- "Std"
int1.name <- 'Internal'
sam1.name <- 'Sample'
blank.name <- 'Blank'
expblank.name <- 'Sample blank'

# Internal check info
int.ppm <- 25 # expected concentration of internal checks in ppm

# Sample dilution info
dil <- 23 # hand-dilution of samples before running

# External samples aren't helpful here because TN wasn't measured
```

## Step 1: Import data and subset standards, checks, samples
```{r, echo=FALSE}
# Import raw data with header notes removed manually in excel or text editor
raw = read.csv("~/SFAgrowthrate/SFAphase1/DOC/SFA1_DOC_52-102.csv", fileEncoding = "UTF-8-BOM")

# Remove excluded injections
raw <- raw[which(raw$Excluded == 0),]

# Fix column names
names(raw)[names(raw) == 'Date...Time'] <- 'DT'
names(raw)[names(raw) == 'Sample.Name'] <- 's.type'
names(raw)[names(raw) == 'Analysis.Inj..'] <- 'analys.type'

# Remove junk columns
raw <- raw[,c(1,3:4,9,12:14)]

# Subset standard data and remove unnecessary columns
stds <- raw[which(raw$s.type ==stds.name),]

# Subset internal checks
ints <- raw[which(raw$s.type == int1.name),]

# Subset samples
samples <- raw[which(raw$s.type == sam1.name | raw$s.type == expblank.name | raw$s.type == blank.name),]
                       
```

## Step 2: Assess standard curve
```{r, results="show"}
# NPOC
npoc.lm <- lm(Area ~ Conc., data = stds)
plot(Area ~ Conc., data = stds, main = "Initial Cal Curve NPOC") +
  abline(npoc.lm)
summary(npoc.lm)
cal.NPOC <- as.data.frame(tidy(npoc.lm))
```

## Step 3: Assess external checks

Not necessary because did not measure TN.

## Step 4: Assess internal checks for drift over time
```{r, echo=FALSE}
# Convert date/time to POSIXct (format= specifies input format)
ints$DT.new = as.POSIXct(ints$DT, format="%m/%d/%Y %H:%M")

# Time of first check
t1 = ints[1,8]

# Create time elapsed column
ints$seconds = round((ints$DT.new - t1), 2)
ints$minutes = as.numeric(ints$seconds/60)

# Summarize mean time elapsed of each internal check group
int.times <- ints %>% group_by(Sample.ID) %>% summarize(mean.min = mean(minutes))

# Calculate concentration in internal standards based on curve
# I've found that the concentrations that the machine comes up with have no relationship to the standard curve for some reason
ints <- ints %>% mutate(Corr.Conc.=(Area-cal.NPOC[1,2])/(cal.NPOC[2,2]))
int.conc <- ints %>% group_by(analys.type, Sample.ID) %>% summarize(mean.conc = mean(Corr.Conc.))
ints.summary <- inner_join(int.times, int.conc, by='Sample.ID')

# Calculate concentration of NPOC in standards based on standard curve
NPOC.stds.summary <- stds %>% group_by(Conc.) %>% summarize(mean.area = mean(Area))
NPOC.stds.summary <- NPOC.stds.summary %>% mutate(calc.conc = (mean.area-cal.NPOC[1,2])/cal.NPOC[2,2])

# Actual concentrations expected from internal checks based on standard curves
NPOC.std.25ppm <- NPOC.stds.summary[NPOC.stds.summary$Conc.==25,]$calc.conc

# Calculate percent and proportion drift
ints.summary <- ints.summary %>% mutate(percent.drift = ((mean.conc-NPOC.std.25ppm)/NPOC.std.25ppm)*100)
ints.summary <- ints.summary %>% mutate(prop.drift = (mean.conc/NPOC.std.25ppm))

# Plot internal standards over time
ggplot(ints.summary, aes(mean.min, percent.drift, colour = analys.type, fill = analys.type)) + 
  geom_point() +
  labs(x = "Time (min)", y = "% Drift from Original Curve Value") +
  ggtitle("Check Standard Drift - Percentage Basis") + 
    theme_bw() + theme(panel.grid.major = element_blank(), 
                       panel.grid.minor = element_blank()) +
  geom_smooth(method='lm')

ggplot(ints.summary, aes(mean.min, prop.drift, colour = analys.type, fill = analys.type)) + 
  geom_point() +
  labs(x = "Time (min)", y = "Proportion of Original Value") +
  ggtitle("Check Standard Drift - Proportional Basis") + 
    theme_bw() + theme(panel.grid.major = element_blank(), 
                       panel.grid.minor = element_blank()) +
  geom_smooth(method='lm')

```

Drift is definitely there. Will perform drift correction. Drift is somewhat erratic so I will use the overall drift instead of time intervals.

## Step 5: Correct drift and recalculate samples based on standard curve

```{r}
# NPOC drift regression
drift.fix <- lm(prop.drift ~ mean.min, data=ints.summary)
drift.fix <- tidy(drift.fix)
drift.fix <- as.data.frame(drift.fix)

# Re-calculate sample concentrations based on std curve
samples <- samples %>% mutate(Corr.Conc.=(Area-cal.NPOC[1,2])/(cal.NPOC[2,2]))

# Convert date/time to POSIXct
samples$DT.new = as.POSIXct(samples$DT, format="%m/%d/%Y %H:%M")

# Create time elapsed (minutes) since first internal check
samples$sec.elapsed = round((samples$DT.new - t1), 2)
samples$min.elapsed = as.numeric(samples$sec.elapsed/60)
samples$min.elapsed = as.numeric(samples$min.elapsed)

# NPOC drift correction
samples <- mutate(samples, drift.conc = Corr.Conc./(min.elapsed*drift.fix[2,2]+drift.fix[1,2]))

# Correct for hand-dilution (not dilutions done by Shimadzu, which are already accounted for)
samples.final <- samples %>% mutate(FinalConc. = drift.conc*dil) 

# Remove blanks from dataset
samples.final <- samples.final[!samples.final$s.type=="Blank",]

# Summarize data (average injections of same sample)
samples.final.summary <- samples.final %>% group_by(s.type, Sample.ID) %>% summarize(mean = mean(FinalConc.), stdv=sd(FinalConc.))

samples.final.summary$mean <- round(samples.final.summary$mean, 2)
samples.final.summary

# Save output for further analysis
write.csv(samples.final[,c(2,3,13)], file ="~/SFAgrowthrate/SFAphase1/DOC/SFA1_DOC_52-102_processed.csv", row.names = FALSE)
write.csv(samples.final.summary, file ="~/SFAgrowthrate/SFAphase1/DOC/SFA1_DOC_52-102_processed_summary.csv", row.names = FALSE)

```