---
title: "SFA2 - CO2 analysis"
author: "Cassandra Wattenburger"
date: "12/7/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

**Goal:** Analyze headspace CO2 from microcosms.


# Import libraries, clear working directory

```{r}
library(tidyverse)
library(cowplot)

sessionInfo()

rm(list=ls())
```



# Import GCMS data

Note: Exported from PostRun Analysis software as ASCII

This script is built to run on data that has been stored in a particular project directory containing a "data" directory containing sub-directories for each time point for which data was collected, containing .txt files with the GCMS output for each sample.

```{r}
# List of directories containing data files
dir = list.dirs(path="data_CO2/SFA2/data", full.names=TRUE, recursive=FALSE)

# Create lists of .txt files from directories
files = list()
for (i in dir) {
  files[[i]] <- list.files(path=i, pattern="*.txt", full.names=TRUE, recursive=FALSE)
}

# Extract area data
# TIC, w/z44 (12C), w/z45 (13C)

# Construct data frame containing all data
area.df = tibble() # initialize empty tibble

for (i in names(files)) { # for each list of files (refers to a directory)
  for (x in files[[i]]) { # for each file in the selected list (directory)
    # Create sample names based on file names
    # Strips the rest of the path from the sample name at the end
    name = str_remove(x, "data_CO2/SFA2/data/.+/") %>%
      str_remove(".txt")
    #name.list[[i]] = append(name.list[[i]], name)
    
    # Isolate time point info
    date = str_remove(x, "data_CO2/SFA2/data/") %>%
      str_remove("/.+.txt") %>%
      str_remove(".+_")
    tp = str_remove(x, "data_CO2/SFA2/data/") %>%
      str_remove("/.+.txt") %>%
      str_remove("_.+")
    
    # Isolate the area data from the raw data files and clean up
    dat = read_tsv(x, skip=7, col_names=TRUE) %>%
      select("Measurement" = Name, Area) %>% # keep only the name of sample and the area measurements collected
      drop_na() %>%
      mutate(Measurement = case_when(Measurement == "TIC" ~ "Ctotal", # more explicit measurement labels
                              Measurement == "m/z 44" ~ "C12",
                              Measurement == "m/z 45" ~ "C13"),
             TP = tp, # adding additional metadata...
             Date = date,
             Sample = name)
    area.df = rbind(area.df, dat) # add to master data frame
  }
}
```


# Calculate total area

TIC area often glitches and does not record (not sure why), so we will use 12C + 13C instead which is a close approximation.

```{r}
# Calculate
area.df = area.df %>%
  pivot_wider(names_from = Measurement, values_from = Area) %>%
  mutate(C1213 = C12+C13)
```


# Standards

Separate from samples/atmosphere measurements

```{r}
std.df = area.df %>%
  filter(str_detect(Sample, "std"))
```

Correct split-ratios and add ppm of each standard

* standards 7-9 were run at a split-ratio of 90% rather than 50% because of the high CO2 concentration
* ppm calculated based on volume and % CO2 used and bottle volume

Multiplier: 0.5x = 0.9, x = 1.8

```{r}
# Calculate adjusted areas
std.df = std.df %>%
  mutate(Area.adj = if_else(str_detect(Sample, "[7-9]"), C1213*1.8, C1213, missing=NULL), # correct split-ratio
         ppm = case_when(Sample == "std0" ~ 0, # add ppm
                         Sample == "std1" ~ 769,
                         Sample == "std2" ~ 1538,
                         Sample == "std3" ~ 3077,
                         Sample == "std4" ~ 7692,
                         Sample == "std5" ~ 15385,
                         Sample == "std6" ~ 30769,
                         Sample == "std7" ~ 76923,
                         Sample == "std7.5" ~ 153856,
                         Sample == "std8" ~ 230769,
                         Sample == "std8.5" ~ 307692,
                         Sample == "std9" ~ 384615)) 
  
```

Visualize

```{r}
std.df %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Date, scales="free") +
  theme_test()
```

Calibration curve

```{r}
# Linear regressions
std.cal = tibble()
for (x in unique(std.df$Date)) {
  dat = filter(std.df, Date==x)
  std.lm = lm(Area.adj ~ ppm, dat)
  this.row = bind_cols(coeff = summary(std.lm)$coefficients[2,1], adjR2 = summary(std.lm)$adj.r.squared , pval = summary(std.lm)$coefficients[2,4]) # save coefficent, adj. R2, and p-value
  std.cal = bind_rows(std.cal, this.row)
}

std.cal
```


Remove std 1 - I think it would be better to include another high ppm sample instead of lower ones.

```{r}
std.df %>%
  filter(Sample != "std1") %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Date, scales="free") +
  theme_test()
```

Calibration curve - no std1

```{r}
# Linear regressions
std.cal = tibble()
for (x in unique(std.df$Date)) {
  dat = filter(std.df, Date==x & Sample != "std1")
  std.lm = lm(Area.adj ~ ppm, dat)
  this.row = bind_cols(coeff = summary(std.lm)$coefficients[2,1], adjR2 = summary(std.lm)$adj.r.squared , pval = summary(std.lm)$coefficients[2,4]) # save coefficent, adj. R2, and p-value
  std.cal = bind_rows(std.cal, this.row)
}

std.cal
```


Remove std 1 and 4

```{r}
std.df %>%
  filter(!(Sample %in% c("std1", "std4"))) %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Date, scales="free") +
  theme_test()
```

Calibration curve - no std1 or std4

```{r}
# Linear regressions
std.cal = tibble()
for (x in unique(std.df$Date)) {
  dat = filter(std.df, Date==x & Sample != "std1")
  std.lm = lm(Area.adj ~ ppm, dat)
  this.row = bind_cols(coeff = summary(std.lm)$coefficients[2,1], adjR2 = summary(std.lm)$adj.r.squared , pval = summary(std.lm)$coefficients[2,4]) # save coefficent, adj. R2, and p-value
  std.cal = bind_rows(std.cal, this.row)
}

std.cal
```

Same result. Will add more higher ppm standards and omit stds 1 and 4 for future runs.


# Samples

Isolate microcosm and atmosphere samples

I just want to make sure that the raw sample areas aren't exceeding the standards.

```{r}
sample.df = anti_join(area.df, std.df) %>%
  mutate(ppm = 0) # dummy ppm
```

Plot against standards:

```{r}
sample.df %>%
  ggplot(aes(y=C1213, x=TP)) +
  scale_y_continuous(limits=c(0,5000000)) +
  labs(x="50% split", t="Area") +
  geom_boxplot() +
  theme_test() +
  theme(panel.grid = element_blank())

# stds
std.df %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point(shape=1) +
  geom_point(aes(y=C1213, x=ppm, color=Date), data=sample.df) +
  #scale_y_continuous(limits=c(0,500000)) +
  #scale_x_continuous(limits=c(0, 20000)) +
  theme_test() +
  theme()
```

Might consider adding a std 10 that is 6 or 7 mL 100% CO2 as well...