---
title: "SFA2 - prerocessing"
author: "Cassandra Wattenburger"
date: "9/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = "show")
knitr::opts_chunk$set(message = FALSE)
```

```{r}
# Clear working directory, load in packages, generate package info
rm(list=ls())

library("tidyverse")
library("ape")
library("phyloseq")

sessionInfo()
```

### Import data

```{r}
# Rep 1
count_rep1 <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/rep1/final/SFA2_rep1.counts-final.tsv")
tax_rep1 <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/rep1/final/SFA2_rep1.taxonomy-final.tsv")

# Rep 2
count_rep2 <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/rep2/final/SFA2_rep2.counts-final.tsv")
tax_rep2 <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/rep2/final/SFA2_rep2.taxonomy-final.tsv")

# Rep 3
count_rep3 <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/rep3/final/SFA2_rep3.counts-final.tsv")
tax_rep3 <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/rep3/final/SFA2_rep3.taxonomy-final.tsv")

# Redos
count_redos <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/redos/final/SFA2_redos.counts-final.tsv")
tax_redos <- read_tsv(file="~/SFAgrowthrate/data_amplicon/SFA2_full/redos/final/SFA2_redos.taxonomy-final.tsv")

# Tree
tree <- read.tree(file="~/SFAgrowthrate/data_amplicon/SFA2_full/tree/SFA2_full.tree-final.nwk")

# Metadata
meta <- read_tsv(file = "~/SFAgrowthrate/data_amplicon/SFA2_full/SFA2_full_metadata.tsv")
```

Combine replicate and redo data

```{r}
# Count data combine
count <- full_join(count_rep1, count_rep2, by="ASV") %>%
  full_join(count_rep3, by="ASV") %>% 
  full_join(count_redos, by="ASV")

# Taxonomy table combine
tax <- full_join(tax_rep1, tax_rep2) %>%
  full_join(tax_rep3) %>% 
  full_join(tax_redos)
```

### Import into phyloseq

```{r}
# Reformat and import count table
count_matrix <- count %>%
  column_to_rownames(var="ASV") %>%
  rename_with(~ gsub("(.+)", "sa\\1", .x)) %>%
  as.matrix()

phy_count <- otu_table(count_matrix, taxa_are_rows=TRUE)

# Reformat and import taxonomy table
tax_matrix <- tax %>% 
  column_to_rownames(var="ASV") %>% 
  as.matrix()

phy_tax <- tax_table(tax_matrix)

# Metadata import
phy_meta <- sample_data(meta)

# Create phyloseq object
physeq <- phyloseq(phy_count, phy_tax, phy_meta, tree)

physeq
```

### Remove non-prokaryotic ASVs

Mitochondrial, chloroplast, protist, unknown domain sequences.

```{r}
# Remove non-prokaryotic ASVs
physeq_remove_asvs <- taxa_names(subset_taxa(physeq, Order=="Chloroplast" | Family=="Mitochondria" |
                                          Domain=="Unassigned" | Domain=="putative Unassigned" |
                                          Domain=="putative Bacteria"))
physeq_all_asvs <- taxa_names(physeq)
physeq_keep_asvs <- physeq_all_asvs[!(physeq_all_asvs %in% physeq_remove_asvs)]

physeq2 <- prune_taxa(physeq_keep_asvs, physeq)
physeq2
```

### Sparcity filters

Keep only ASVs that occur in at least three samples.

```{r}
# Filter out ASVs that appear in less than three samples
sparse_pass_asvs <- count %>%
  mutate(across(-ASV, ~if_else(.x > 0, 1, 0), {.cols})) %>%
  mutate(total = rowSums(.[-1])) %>%
  filter(total > 2) %>%
  .$ASV

physeq3 <- prune_taxa(sparse_pass_asvs, physeq2)
physeq3
```

### Read depth

```{r}
# Extract count and metadata from processed phyloseq object
count_processed <- data.frame(otu_table(physeq3))
tax_processed <- data.frame(tax_table(physeq3))
meta_processed <- data.frame(sample_data(physeq3)) %>%
  select(-SampleID) %>%
  rownames_to_column(var="SampleID")

# Calculate read depth per sample
read_depth <- count_processed %>%
  rownames_to_column(var="ASV") %>%
  gather(SampleID, count, -ASV) %>%
  spread(ASV, count) %>%
  mutate(total_reads = rowSums(.[-1])) %>%
  select(SampleID, total_reads) %>%
  inner_join(meta_processed) %>% 
  mutate(Attempt = as_factor(Attempt)) %>% 
  as_tibble()
  
# Visualize
read_depth %>%
  mutate(Replicate = as_factor(Replicate)) %>%
  filter(Type != "pcr_control") %>%
  ggplot(aes(x=reorder(Sample, -total_reads), y=total_reads, fill=Replicate)) +
  geom_col() +
  labs(x="Sample", y="Reads") +
  theme_test()
```

### Compare original to redo samples

Compare sequence depth, Aquifex spike-in proportion, ASV richness.

```{r}
# Isolate resequenced samples and their original counterparts
reseq_sams <- filter(read_depth, Attempt==2) %>% 
  .$Sample
read_depth_reseq <- filter(read_depth, Sample %in% reseq_sams)

# Isolate Aquifex ASV
aquifex <- tax_processed %>%
  filter(Genus == "Aquifex") %>%
  rownames_to_column(var="ASV") %>%
  select(ASV) %>%
  .$ASV

# Spike-in read counts
spike_counts <- count_processed %>%
  rownames_to_column(var="ASV") %>% 
  filter(ASV == "445e8681c3c1a735760e6c394f5f4d0a") %>% 
  select(-ASV) %>%
  gather(variable, value) %>% 
  select(SampleID = variable, std_count = value)

# ASV richness
richness <- count_processed %>% 
  mutate(across(everything(), ~ if_else(.x > 0, 1, 0), {.cols})) %>%
  colSums() %>%
  data.frame() %>% 
  rownames_to_column(var="SampleID")

# Combine
reseq_compare <- inner_join(read_depth_reseq, spike_counts) %>% 
  inner_join(richness) %>% 
  select(c(SampleID, Sample, Attempt:PCR_plate, total_reads, std_count, richness=`.`)) %>% 
  mutate(Replicate = as_factor(Replicate))
```

Visualize:

```{r}
# Read depth
reseq_compare %>% 
  filter(Type != 'PCR control') %>% 
  ggplot(aes(x=Sample, y=total_reads, fill=Attempt, color=Attempt)) +
  geom_col(position = "dodge") +
  theme_test()

# Spike-in counts
reseq_compare %>% 
  filter(Type != 'PCR control') %>% 
  ggplot(aes(x=Sample, y=std_count, fill=Attempt, color=Attempt)) +
  geom_col(position = "dodge") +
  theme_test()

reseq_compare %>% 
  mutate(std_prop = std_count/(total_reads + std_count)) %>% 
  filter(Type != 'PCR control') %>% 
  ggplot(aes(x=Sample, y=std_prop, fill=Attempt, color=Attempt)) +
  geom_col(position = "dodge") +
  theme_test()

# Richness
reseq_compare %>% 
  filter(Type != 'PCR control') %>% 
  ggplot(aes(x=Sample, y=richness, fill=Attempt, color=Attempt)) +
  geom_col(position = "dodge") +
  theme_test()
```

Attempt 2 seems to be improvement in most cases.

Isolate cases in which original has higher read depth, richness, std count, but equal or lower std proportion:

```{r}
# Reformat
reseq_compare_wide <- reseq_compare %>% 
  select(-c(SampleID, PCR_plate, Library)) %>% 
  mutate(std_prop = std_count/(total_reads+std_count)) %>% 
  pivot_wider(names_from = Attempt, names_glue = "{.value}_{Attempt}", values_from = c(total_reads, richness, std_count, std_prop))

# Comparison
reseq_compare_wide %>% 
  mutate(keep = if_else(total_reads_1 > total_reads_2 & richness_1 > richness_2 & std_count_1 > std_count_2 & std_prop_1 <= std_prop_2, 1, 2))
```

The redo sample is always better.

Replace originals with redos in dataset:

```{r}
# Isolate original samples
rm_samIDs <- filter(reseq_compare, Attempt==1) %>%
  .$SampleID %>% 
  gsub("sa([0-9]+)", "\\1", .)
  
# Create "not in" operator
`%!in%` <- Negate(`%in%`)

# Remove from phyloseq object
physeq4 <- subset_samples(physeq3, SampleID %!in% rm_samIDs)
physeq4
```

### Normalize to the internal standard

For each taxa in each sample, taxa_counts/std_counts. Not rarefying.

```{r}
# Extract count from phyloseq object
count_reseq <- data.frame(otu_table(physeq4))

# Normalize
count_normalized <- count_reseq %>%
  rownames_to_column(var="ASV") %>%
  gather(SampleID, value, -ASV) %>%
  spread(ASV, value) %>%
  mutate(across(-SampleID, ~ .x/`445e8681c3c1a735760e6c394f5f4d0a`, {.col})) # Divide ASV counts by Aquifex ASV counts
```

Save intermediate data:

```{r, eval=FALSE}
# ASV counts
count_normalized2 <- count_normalized %>% 
  gather(ASV, value, -SampleID) %>% 
  spread(SampleID, value)
  
saveRDS(count_normalized2, file="../data_intermediate/SFA2_count_normalized.rds")

# Taxonomy
tax_reseq <- data.frame(tax_table(physeq4))
saveRDS(tax_reseq, file="../data_intermediate/SFA2_tax_processed.rds")

# Tree
tree_reseq <- phy_tree(physeq4)
saveRDS(tree_reseq, file="../data_intermediate/SFA2_tree_processed.rds")
```





