---
title: "SFA2 - CO2 analysis"
author: "Cassandra Wattenburger"
date: "12/7/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

**Goal:** Analyze headspace CO2 from microcosms.


# Import libraries, clear working directory

```{r}
library(tidyverse)
library(cowplot)

sessionInfo()

rm(list=ls())
```


# Import and reformat GCMS data

Note: Exported from Shimadzu PostRun Analysis software as ASCII using second and fourth output options.

This script is built to run on data that has been stored in a project directory containing a "data" directory with sub-directories for each time point that .txt files with the GCMS output for each sample.

```{r}
# List of directories containing data files
dir = list.dirs(path="../data_CO2/SFA2/data", full.names=TRUE, recursive=FALSE)

# Create lists of .txt files from directories
files = list()
for (i in dir) {
  files[[i]] <- list.files(path=i, pattern="*.txt", full.names=TRUE, recursive=FALSE)
}

# Extract area data
# TIC, w/z44 (12C), w/z45 (13C)

# Construct data frame containing all data
area.df = tibble() # initialize empty tibble

for (i in names(files)) { # for each list of files (refers to a directory)
  for (x in files[[i]]) { # for each file in the selected list (directory)
    # Create sample names based on file names
    # Strips the rest of the path from the sample name at the end
    name = str_remove(x, "../data_CO2/SFA2/data/.+/") %>%
      str_remove(".txt")
    #name.list[[i]] = append(name.list[[i]], name)
    
    # Isolate time point info
    date = str_remove(x, "../data_CO2/SFA2/data/") %>%
      str_remove("/.+.txt") %>%
      str_remove(".+_")
    tp = str_remove(x, "../data_CO2/SFA2/data/") %>%
      str_remove("/.+.txt") %>%
      str_remove("_.+")
    
    # Isolate the area data and clean up
    dat = read_tsv(x, skip=7, col_names=TRUE) %>%
      select("Measurement" = Name, Area) %>% # keep only the name of sample and the area measurements collected
      drop_na() %>%
      mutate(Measurement = case_when(Measurement == "TIC" ~ "Ctotal", # more explicit measurement labels
                              Measurement == "m/z 44" ~ "C12",
                              Measurement == "m/z 45" ~ "C13"),
             TP = tp, # adding additional metadata...
             Date = date,
             Sample = name)
    area.df = rbind(area.df, dat) # add to master data frame
  }
}

# Parse dates
area.df = area.df %>%
  mutate(Date = gsub("([0-2]{2})([0-9]{2})([0-2]{2})", "\\1-\\2-20\\3", Date)) %>%
  mutate(Date = parse_date(Date, "%m-%d-%Y"))
```

# Calculate total area

TIC area often does not record (not sure why), so we will use 12C + 13C instead which is a close approximation.

```{r}
# Calculate
area.df = area.df %>%
  pivot_wider(names_from = Measurement, values_from = Area) %>%
  mutate(C1213 = C12+C13)
```

# Standards

Separate from samples/atmosphere measurements

```{r}
std.df = area.df %>%
  filter(str_detect(Sample, "std"))
```

Correct split-ratios and add ppm of each standard

* standards 7-9 were run at a split-ratio of 90% rather than 50% because of the high CO2 concentration
* ppm calculated based on volume and % CO2 used and bottle volume

Multiplier: 0.5x = 0.9, x = 1.8

```{r}
# Calculate adjusted areas
std.df = std.df %>%
  mutate(Area.adj = if_else(str_detect(Sample, "[7-9]"), C1213*1.8, C1213, missing=NULL), # correct split-ratio
         ppm = case_when(Sample == "std0" ~ 0, # add ppm
                         Sample == "std1" ~ 769,
                         Sample == "std2" ~ 1538,
                         Sample == "std3" ~ 3077,
                         Sample == "std4" ~ 7692,
                         Sample == "std5" ~ 15385,
                         Sample == "std6" ~ 30769,
                         Sample == "std7" ~ 76923,
                         Sample == "std7.5" ~ 153856,
                         Sample == "std8" ~ 230769,
                         Sample == "std8.5" ~ 307692,
                         Sample == "std9" ~ 384615)) 
  
```

Visualize

```{r}
std.df %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Date, scales="free") +
  theme_test()
```

Calibration curve

```{r}
# Linear regressions
std.cal = tibble()
for (x in unique(std.df$Date)) {
  dat = filter(std.df, Date==x)
  std.lm = lm(Area.adj ~ ppm, dat)
  this.row = bind_cols(Date = unique(dat$Date), # save results
                       intercept = summary(std.lm)$coefficients[1,1],
                       slope = summary(std.lm)$coefficients[2,1], 
                       adjR2 = summary(std.lm)$adj.r.squared, 
                       pval = summary(std.lm)$coefficients[2,4])
  std.cal = bind_rows(std.cal, this.row)
}

std.cal
```

Remove std 1 - I think it would be better to include another high ppm sample instead of lower ones.

```{r}
std.df %>%
  filter(Sample != "std1") %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Date, scales="free") +
  theme_test()
```

Calibration curve - no std1

```{r}
# Linear regressions
std.cal1 = tibble()
for (x in unique(std.df$Date)) {
  dat1 = filter(std.df, Date==x & Sample != "std1")
  std.lm1 = lm(Area.adj ~ ppm, dat)
  date1 = unique(dat1$Date)
  this.row1 = bind_cols(Date = date, coeff = summary(std.lm1)$coefficients[2,1], adjR2 = summary(std.lm1)$adj.r.squared , pval = summary(std.lm1)$coefficients[2,4]) # save coefficent, adj. R2, and p-value
  std.cal1 = bind_rows(std.cal, this.row)
}

std.cal1
```

Remove std 1 and 4

```{r}
std.df %>%
  filter(!(Sample %in% c("std1", "std4"))) %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Date, scales="free") +
  theme_test()
```

Calibration curve - no std1 or std4

```{r}
# Linear regressions
std.cal14 = tibble()
for (x in unique(std.df$Date)) {
  dat14 = filter(std.df, Date==x & Sample != "std1")
  std.lm14 = lm(Area.adj ~ ppm, dat14)
  this.row14 = bind_cols(coeff = summary(std.lm14)$coefficients[2,1], adjR2 = summary(std.lm14)$adj.r.squared , pval = summary(std.lm14)$coefficients[2,4]) # save coefficent, adj. R2, and p-value
  std.cal14 = bind_rows(std.cal14, this.row14)
}

std.cal14
```

Same result. Will add more high ppm standards and omit stds 1 and 4 for future runs.

# Samples

Isolate microcosm samples

```{r}
sample.df = anti_join(area.df, std.df) %>%
  mutate(dummy_ppm = 0) %>% # dummy ppm for graphing with standards on ppm axis
  filter(!str_detect(Sample, "[a-z]")) %>% # exclude other non-samples 
  mutate(Sample, Sample = parse_number(Sample))# convert to numeric
```

Add metadata information about inoculants

```{r}
# Import metadata
meta = read_csv("../data_CO2/SFA2/SFA2_headspace_metadata.csv")

# Merge with CO2 data
sample.df = full_join(sample.df, meta, by="Sample")
```

Visualize

```{r}
# Compare samples against range of standards
std.df %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point(shape=1) +
  geom_point(aes(y=C1213, x=dummy_ppm, color=Date), data=sample.df) +
  labs(title="Samples vs. standards") +
  theme_test()

# Observe progression of CO2 production
sample.df %>%
  mutate(Control = if_else(Sample %in% c(379:384), "Y", "N")) %>% # distinguish negative control
  ggplot(aes(y=C1213, x=Date)) +
  labs(y="Area") +
  geom_jitter(aes(color = Control)) +
  labs(title="Quick look - respiration over time") +
  theme_test()
```

Outlier ~ Dec. 15, looking at lab notes it seems I may have accidentally sampled twice into the same vial for this sample. This means this value is ~2x more than what it should be. I'll correct this by halving the area value to approximate the correct measurement.

```{r}
# Isolate outlier and correct
outlier = sample.df %>%
  subset(Sample==363 & TP=="tp6") %>%
  mutate(Ctotal = Ctotal/2, C12 = C12/2, C13 = C13/2, C1213 = C1213/2) # halve measurements

# Replace in dataframe
sample.df = sample.df %>%
  filter(!(Sample==363 & TP=="tp6")) %>%
  bind_rows(outlier)
```

### Calculate CO2

Use calibration curve to convert sample areas into ppm.

```{r}
# Convert sample area to ppm using calibration curves
sample.conv = tibble()
for (d in unique(sample.df$Date)) {
  cal = std.cal %>% filter(Date==d)
  conv = sample.df %>% 
    filter(Date==d) %>%
    mutate(ppm = (C1213 - cal$intercept)/cal$slope)
  sample.conv = rbind(sample.conv, conv)
}
```

Sanity check visualization

```{r}
# Sanity check, plot against standards
std.df %>%
  ggplot(aes(x=ppm, y=Area.adj)) +
  geom_point(shape=1) +
  geom_point(aes(y=C1213, x=ppm, color=Date), data=sample.conv) +
  #scale_y_continuous(limits=c(0,500000)) +
  #scale_x_continuous(limits=c(0, 20000)) +
  theme_test() +
  theme()
```

Calculate mol CO2

```{r}
# Ideal gas law: n=(P*V)/(R*temp)
# NOTE: pressure does not matter for this calulation because it's all relative to the stds
R = 0.082057338 # universal gas constant, (L*atm)/(mol*K)
temp = 296.48 # K, 74F in lab throughout experiment
V = 0.013 # L, headspace of 30 mL serum bottle with sand ########### NEED TO DBL CHECK##########
mwCO2 = 44.01 # molecular weight (g/mol) of CO2
P = 1 #atm, assumes 1 atm in ucosms during sampling

# Calculate per time point
molCO2100 = P*V/(R*temp) # total mol of CO2 in headspace, if it were 100% CO2
mgCO2100 = molCO2100 * mwCO2*1000 # mol * g/mol * 1000mg/g = mg, mg of CO2 present if it were 100% CO2
sample.conv = mutate(sample.conv, mgCO2 = mgCO2100*(ppm*10^-6)) # actual mg CO2 based on ppm

# Cumulative mg CO2 and C
sample.totals = sample.conv %>%
  group_by(Sample, Inoculant, DOC_predict) %>%
  summarize(totalCO2mg = sum(mgCO2), totalCmg = totalCO2mg*(12/44)) %>% 
  ungroup()
```

Visualize

```{r}
sample.totals %>%
  ggplot(aes(x=Inoculant, y=totalCO2mg, color=DOC_predict)) +
  geom_point() +
  labs(title="Cumulative CO2 respiration", y="CO2 mg", x="Inoculant") +
  theme_test()

resp_plot <- sample.totals %>%
  filter(DOC_predict %in% c("high", "low")) %>% 
  ggplot(aes(x=DOC_predict, y=totalCO2mg)) +
  geom_boxplot() +
  labs(y="Cumulative CO2 (mg)", x="DOC \"phenotype\"") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=16),
        axis.text = element_text(size=14))
resp_plot
```

```{r, eval=FALSE}
# BOOKMARK
saveRDS(sample.totals, file="../data_CO2/SFA2/SFA2_CO2.rds")
```

# CO2 respiration over time

Are there treatment effects in different phases of the incubation?

Early = tp 3-7
Late = tp 8+

Based on earlier visual of respiration over time, seems to be a shift in activity after tp 7.

```{r}
# Split into early and late
sample.early = subset(sample.conv, TP %in% c("tp0","tp1","tp2","tp3","tp4","tp5","tp6","tp7"))
sample.late = anti_join(sample.conv, sample.early)

# Calculate cumulative CO2 for early and late
sample.total.early = sample.early %>%
  group_by(Sample, Inoculant, DOC_predict) %>%
  summarize(totalCO2mg = sum(mgCO2), totalCmg = totalCO2mg*(12/44)) %>%
  add_column(Incubation_stage = "early") %>% 
  ungroup()

sample.total.late = sample.late %>%
  group_by(Sample, Inoculant, DOC_predict) %>%
  summarize(totalCO2mg = sum(mgCO2), totalCmg = totalCO2mg*(12/44)) %>%
  add_column(Incubation_stage = "late") %>% 
  ungroup()

sample.total.split = bind_rows(sample.total.early, sample.total.late)
```

Visualize

```{r}
sample.total.split %>%
  ggplot(aes(x=Incubation_stage, y=totalCO2mg, color=DOC_predict)) +
  geom_boxplot() +
  theme_test()
```

