---
title: "SFA - CUE vs community dissimilarity"
author: "Cassandra Wattenburger"
date: "11/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
# Import libraries
library(phyloseq)
library(tidyverse)
library(cowplot)

sessionInfo()

# Clear environment
rm(list=ls())
```

# Import data

* metadata
* growth efficiency calculations
* rarefied physeq
* unrarefied physeq

```{r}
# Metadata
meta <- read_tsv("../data_amplicon/SFA2_full/SFA2_metadata_corrected.tsv")
# Growth efficiency
groweff <- readRDS("../data_intermediate/SFA2_growtheff.rds")
# Rarefied count
physeq_rare <- readRDS("../data_intermediate/SFA2_physeq_rare.rds")
# Not rarefied counts
physeq_unrare <- readRDS("../data_intermediate/SFA2_physeq_unrare.rds")
```

# Rarefying losses

How much data lost to rarefication?

```{r}
# How many taxa/samples are lost to rarefication?
physeq_rare
physeq_unrare

# Per sample?
rare_count <- data.frame(otu_table(physeq_rare))
unrare_count <- data.frame(otu_table(physeq_unrare))

# Count ASVs per sample
rare_binary <- rare_count %>% 
  mutate(across(everything(), ~if_else(. > 1, 1, 0)))
rare_binary <- colSums(rare_binary)
rare_sams <- names(rare_binary) # samples left after rarefication

unrare_binary <- unrare_count %>% 
  mutate(across(everything(), ~if_else(. > 1, 1, 0)))
unrare_binary <- colSums(unrare_binary)
unrare_binary <- unrare_binary[names(unrare_binary) %in% rare_sams] # filter samples not in rarefied
  
# Combine
rich_compare <- data.frame(rare=rare_binary, unrare=unrare_binary) %>% 
  mutate(diff = unrare - rare)

# Visualize
rich_compare %>% 
  ggplot(aes(x=reorder(rownames(.), diff), y=diff)) +
  geom_point() +
  labs(x="sample", y="difference in ASVs") +
  theme_test()

rich_compare %>%
  mutate(SampleID = rownames(.),
         SampleID = gsub("sa", "", SampleID),
         SampleID = as.numeric(SampleID)) %>% 
  inner_join(meta) %>% 
  ggplot(aes(x=Day, y=diff)) +
  geom_point() +
  labs(x="Day", y="difference in ASVs") +
  theme_test()
```

Calculate relative abundances:

```{r}
physeq_relabund <- transform_sample_counts(physeq_unrare, function(x) x/sum(x))
```

Compare oridnations:

```{r}
set.seed(2)

# Rarefied
wuni_rare <- distance(physeq_rare, method="wunifrac")
rare_pcoa <- ordinate(physeq_rare, method="PCoA", distance=wuni_rare)

# Relative abundances
wuni_relabund <- distance(physeq_relabund, method="wunifrac")
relabund_pcoa <- ordinate(physeq_relabund, method="PCoA", distance=wuni_relabund)

# Visualize
plot_ordination(physeq_rare, rare_pcoa, color="Day") +
  theme_test()
plot_ordination(physeq_relabund, unrare_pcoa, color="Day") +
  theme_test()
```

# Calculate community dissimilarity

* Using relative abundance data because it looks similar to rarefied w/similar dispersion but doesn't throw out data

Average replicates:

```{r}
nongrowth_sams <- filter(meta, Type!="Growth")
nongrowth_sams <- nongrowth_sams$SampleID

# Average technical replicates
relabund_avg <- data.frame(otu_table(physeq_relabund)) %>% # lots of reformatting needed
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="SampleID") %>%  
  pivot_longer(cols=c(-SampleID), names_to="ASV", values_to="relabund") %>% # long format
  mutate(SampleID = gsub("sa", "", SampleID),
         SampleID = as.numeric(SampleID)) %>% 
  filter(!(SampleID %in% nongrowth_sams)) %>% 
  inner_join(meta) %>% # add metadata
  mutate(ucosm = paste0(Inoculant, Day)) %>%
  group_by(ucosm, ASV) %>% 
  summarize(relabund_avg = mean(relabund)) %>% # calc. mean
  ungroup() %>% 
  pivot_wider(names_from=ASV, values_from=relabund_avg) %>% 
  column_to_rownames(var="ucosm") %>% # reformat back
  as.matrix() %>% 
  t()

meta2 <- meta %>% 
  filter(Type=="Growth") %>% 
  mutate(ucosm = paste0(Inoculant, Day)) %>% 
  select(ucosm, Type, Inoculant, Day) %>% 
  unique() %>% 
  column_to_rownames(var="ucosm")

physeq_relabund_avg <- phyloseq(otu_table(relabund_avg, taxa_are_rows = TRUE), 
                                sample_data(meta2), phy_tree(physeq_relabund))

# String sorting function stolen from https://gist.github.com/martinctc/56b3fb701a182f5b8dffceecd65b6d86
str_arrange <- function(x) {
  x %>%
    stringr::str_split("") %>% # Split string into letters
    purrr::map(~sort(.) %>% paste(collapse = "")) %>% # Sort and re-combine
    as_vector() 
  } # Convert list into vector

# Calculate distances
wuni_relabund_avg <- distance(physeq_relabund_avg, method="wunifrac")

# Reformat
wuni_dist_df <- wuni_relabund_avg %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="Sample1") %>% 
  pivot_longer(-Sample1, names_to = "Sample2", values_to = "wuni_dist") %>% 
  filter(wuni_dist!=0) %>% # remove self-comparisons
  mutate(foo = as.character(paste0(Sample1, Sample2)), # remove repeated comparisons
         foo = str_arrange(foo)) %>% 
  bind_cols(dup=duplicated(.$foo)) %>% # id duplicated comparisons
  filter(dup==FALSE) %>% # remove duplicates
  select(Sample1, Sample2, wuni_dist)
```

# Calculate growth eff. diff

```{r}
# Reformat
groweff_mx <- groweff %>% 
  mutate(ucosm = paste0(Inoculant, Day)) %>% 
  filter(!is.na(Day)) %>% # remove NAs
  select(ucosm, growth_eff) %>% 
  column_to_rownames(var="ucosm") %>% 
  as.matrix()

groweff_diff <- dist(groweff_mx, method="maximum") %>% # calculate difference
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="Sample1") %>% 
  pivot_longer(-Sample1, names_to = "Sample2", values_to = "greff_diff") %>% 
  filter(greff_diff!=0) %>% # remove self-comparisons
  mutate(foo = as.character(paste0(Sample1, Sample2)), # remove repeated comparisons
         foo = str_arrange(foo)) %>% 
  bind_cols(dup=duplicated(.$foo)) %>% # id duplicated comparisons
  filter(dup==FALSE) %>% # remove duplicates
  select(Sample1, Sample2, greff_diff)
```

Combine:

```{r}
# Add metadata and combine
meta2 <- rownames_to_column(meta2, var="ucosm")

greff_dist <- inner_join(groweff_diff, wuni_dist_df) %>% 
  inner_join(meta2, by=c("Sample1"="ucosm")) %>% 
  select(Sample1, Inoc1=Inoculant, Sample2, greff_diff, wuni_dist) %>% 
  inner_join(meta2, by=c("Sample2"="ucosm")) %>% 
  select(Sample1, Inoc1, Sample2, Inoc2=Inoculant, Day, greff_diff, wuni_dist)

# Visualize
greff_dist %>% 
  #mutate(color = if_else(Day==30, "yes", "no")) %>% 
  ggplot(aes(x=wuni_dist, y=log(greff_diff))) +
  geom_point(aes(color=Day)) +
  geom_smooth(method="lm", color="black") +
  labs(x="Weighted unifrac, relative abundance normalized", y="ln Difference in growth efficiency") +
  theme_test()
```

Statistics:

Pearson corr

```{r}
cor.test(log(greff_dist$greff_diff), greff_dist$wuni_dist, method="pearson")
```

# Rarefied growth eff diff

Just for comparison.

Average replicates:

```{r}
# Average technical replicates
rare_avg <- data.frame(otu_table(physeq_rare)) %>% # lots of reformatting needed
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="SampleID") %>%  
  pivot_longer(cols=c(-SampleID), names_to="ASV", values_to="rare") %>% # long format
  mutate(SampleID = gsub("sa", "", SampleID),
         SampleID = as.numeric(SampleID)) %>% 
  filter(!(SampleID %in% nongrowth_sams)) %>% 
  inner_join(meta) %>% # add metadata
  mutate(ucosm = paste0(Inoculant, Day)) %>%
  group_by(ucosm, ASV) %>% 
  summarize(rare_avg = mean(rare)) %>% # calc. mean
  ungroup() %>% 
  pivot_wider(names_from=ASV, values_from=rare_avg) %>% 
  column_to_rownames(var="ucosm") %>% # reformat back
  as.matrix() %>% 
  t()

meta2 <- meta %>% 
  filter(Type=="Growth") %>% 
  mutate(ucosm = paste0(Inoculant, Day)) %>% 
  select(ucosm, Type, Inoculant, Day) %>% 
  unique() %>% 
  column_to_rownames(var="ucosm")

physeq_rare_avg <- phyloseq(otu_table(rare_avg, taxa_are_rows = TRUE), 
                                sample_data(meta2), phy_tree(physeq_rare))

# String sorting function stolen from https://gist.github.com/martinctc/56b3fb701a182f5b8dffceecd65b6d86
str_arrange <- function(x) {
  x %>%
    stringr::str_split("") %>% # Split string into letters
    purrr::map(~sort(.) %>% paste(collapse = "")) %>% # Sort and re-combine
    as_vector() 
  } # Convert list into vector

# Calculate distances
wuni_rare_avg <- distance(physeq_rare_avg, method="wunifrac")

# Reformat
wuni_dist_df <- wuni_rare_avg %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="Sample1") %>% 
  pivot_longer(-Sample1, names_to = "Sample2", values_to = "wuni_dist") %>% 
  filter(wuni_dist!=0) %>% # remove self-comparisons
  mutate(foo = as.character(paste0(Sample1, Sample2)), # remove repeated comparisons
         foo = str_arrange(foo)) %>% 
  bind_cols(dup=duplicated(.$foo)) %>% # id duplicated comparisons
  filter(dup==FALSE) %>% # remove duplicates
  select(Sample1, Sample2, wuni_dist)
```

Combine with growth efficiencies:

```{r}
# Add metadata and combine
meta2 <- rownames_to_column(meta2, var="ucosm")

greff_dist <- inner_join(groweff_diff, wuni_dist_df) %>% 
  inner_join(meta2, by=c("Sample1"="ucosm")) %>% 
  select(Sample1, Inoc1=Inoculant, Sample2, greff_diff, wuni_dist) %>% 
  inner_join(meta2, by=c("Sample2"="ucosm")) %>% 
  select(Sample1, Inoc1, Sample2, Inoc2=Inoculant, Day, greff_diff, wuni_dist)

# Visualize
greff_dist %>% 
  #mutate(color = if_else(Day==30, "yes", "no")) %>% 
  ggplot(aes(x=wuni_dist, y=log(greff_diff))) +
  geom_point(aes(color=Day)) +
  geom_smooth(method="lm", color="black") +
  labs(x="Weighted unifrac, rarefied", y="ln Difference in growth efficiency") +
  theme_test()
```

