---
title: "WIP - SFA2 growth estimations exploration"
author: "Cassandra Wattenburger"
date: "11/16/2021"
output: html_document
---

```{r}
# Clear working directory, load in packages, generate package info
rm(list=ls())

library("tidyverse")
library("trelliscopejs")
library("lmerTest")

sessionInfo()
```

# Import data

Files produced in 03_SFA2_growthestimates.Rmd
* growth estimates
* normalized abundance data for time series with estimated growth

```{r}
# Growth
growth <- readRDS("../data_intermediate/SFA2_growth_estimates.rds")

# Taxonomic information
tax <- readRDS("../data_intermediate/SFA2_tax_normalized.rds")
```

# Exploration

### Numbers of growing bacteria

```{r}
# Number of taxa growing
num_grow <- growth %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(num_grew = n()) %>% 
  ungroup()

num_grow

# treatment
num_grow %>% 
  group_by(DOC) %>% 
  summarize(mean_num = mean(num_grew), sd_num = sd(num_grew)) %>% 
  ungroup()

num_grow %>% 
  ggplot(aes(x=DOC, y=num_grew)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Number of growth estimated taxa") +
  theme_test()
```

### Phylogenetic membership:

```{r}
# Merge growth estimates with taxonomy
growth_tax <- inner_join(growth, tax) 

# Graph phylogenetic membership
growth_phylo_plot <- growth_tax %>% 
  group_by(Phylum, DOC, Innoculant) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  arrange(count) %>% 
  ggplot(aes(x=count, y=reorder(Phylum, count), color=DOC)) +
  geom_boxplot() +
  labs(title="Growth", y="", x="Number of ASVs") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=16),
        axis.text = element_text(size=14),
        panel.grid.major.y = element_line(color = "gray",
                                          size = 0.5,
                                          linetype = 2))
growth_phylo_plot
```

```{r, eval=FALSE}
ggsave("../figures/SFA2_growth_phylo.png", growth_phylo_plot, units="mm", width=200, height=100)
```

### Growth parameters:

* specific growth rate (k)
* start day
* end day
* change in abundance
* start abundance
* end abundance

### Sample-level

```{r}
# Summarize parameters
growth_summary <- growth %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(mean_k = mean(k),
            mean_start_day = mean(start_day),
            mean_end_day = mean(end_day),
            mean_end_abund = mean(end_abund),
            total_end_abund = sum(end_abund),
            mean_change_abund = mean(change_abund),
            total_change_abund = sum(change_abund)) %>% 
  ungroup()
```


```{r}
# Mean k
k_mean_plot <- growth_summary %>% 
  ggplot(aes(x=DOC, y=mean_k)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Mean specific growth rate (k)") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=16),
        axis.text = element_text(size=14))
k_mean_plot
  
# Mean start day
lag_mean_plot <- growth_summary %>% 
  ggplot(aes(x=DOC, y=mean_start_day)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Mean lag (days)") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=16),
        axis.text = element_text(size=14))
lag_mean_plot

# Mean change abundance
change_abund_mean_plot <- growth_summary %>% 
  ggplot(aes(x=DOC, y=mean_change_abund)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Mean change in relational abundance") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=14),
        axis.text = element_text(size=14))
change_abund_mean_plot

# Total change abundance
change_abund_total_plot <- growth_summary %>% 
  ggplot(aes(x=DOC, y=total_change_abund)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Total change in relational abundance") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=14),
        axis.text = element_text(size=14))
change_abund_total_plot

# Length of growth
length_plot <- growth_summary %>% 
  mutate(length = mean_end_day - mean_start_day) %>% 
  ggplot(aes(x=DOC, y=length)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Length of growth (days)") +
  theme_test() +
  theme(title = element_text(size=18),
        axis.title = element_text(size=14),
        axis.text = element_text(size=14))
length_plot
```

```{r, eval=FALSE}
ggsave("../figures/SFA2_growth_kmean.png", k_mean_plot, units="mm", width=100, height=100)
ggsave("../figures/SFA2_growth_lagmean.png", lag_mean_plot, units="mm", width=100, height=100)
ggsave("../figures/SFA2_growth_changeabundmean.png", change_abund_mean_plot, units="mm", width=100, height=100)
ggsave("../figures/SFA2_growth_changeabundtotal.png", change_abund_total_plot, units="mm", width=100, height=100)
```

### ASV-level

```{r}
growth_asv <- growth %>% 
  group_by(DOC, ASV) %>% 
  summarize(k = mean(k),
            start_pt = mean(start_pt),
            end_pt = mean(end_pt),
            start_day = mean(start_day),
            end_day = mean(end_day),
            start_abund = mean(start_abund),
            end_abund = mean(end_abund),
            chg_abund = mean(end_abund - start_abund)) %>% 
  ungroup() %>% 
  inner_join(tax)
```

```{r}
growth_asv %>% 
  ggplot(aes(y = Phylum, x=chg_abund)) +
  geom_point() +
  theme_test()
```



### ASV-level distirbutions

Specific growth rates

```{r}
# Distribution of growth rates
growth %>%
  ggplot(aes(x=k, color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="Specific growth rate (k)", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

growth %>%
  ggplot(aes(x=k, color=Innoculant)) +
  geom_density(alpha=0.5) +
  labs(x="Specific growth rate (k)", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

Change in relational abundance

```{r}
# Distribution of change in abundance during growth
growth %>%
  ggplot(aes(x=log(change_abund), color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Change in relational abundance during growth", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

growth %>%
  ggplot(aes(x=log(change_abund), color=Innoculant, linetype=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Change in relational abundance during growth", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

Start relational abundance

```{r}
# Distribution of change in abundance during growth
growth %>%
  ggplot(aes(x=log(start_abund), color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Starting relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

growth %>%
  ggplot(aes(x=log(start_abund), color=Innoculant)) +
  geom_density(alpha=0.5) +
  labs(x="ln Starting in relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

End relational abundance

```{r}
# Distribution of change in abundance during growth
growth %>%
  ggplot(aes(x=log(end_abund), color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Ending relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

growth %>%
  ggplot(aes(x=log(end_abund), color=Innoculant)) +
  geom_density(alpha=0.5) +
  labs(x="ln Ending relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

### Statistics

Specific growth rate:

```{r}
k_lm <- lm(mean_k ~ DOC, data=growth_summary)
hist(resid(k_lm))
plot(predict(k_lm), resid(k_lm))

anova(k_lm)
```

Start day:

```{r}
start_day_lm <- lm(mean_start_day ~ DOC, data=growth_summary)
hist(resid(start_day_lm))
plot(predict(start_day_lm), resid(start_day_lm))

anova(start_day_lm)
```

Mean change in abundance:

```{r}
mean_change_lm <- lm(log(mean_change_abund) ~ DOC, data=growth_summary)
hist(resid(mean_change_lm))
plot(predict(mean_change_lm), resid(mean_change_lm))

anova(mean_change_lm)
```

Total change in abundance:

```{r}
total_change_lm <- lm(total_change_abund ~ DOC, data=growth_summary)
hist(resid(total_change_lm))
plot(predict(total_change_lm), resid(total_change_lm))

anova(total_change_lm)
```

# Relationship to DOC measurement

Import DOC measurements

```{r}
# DOC measurements
doc <- readRDS("../data_DOC/SFA2/SFA2_doc.rds") %>% 
  select(everything(), Sample=sample_id) %>% 
  mutate(Sample = as.numeric(Sample))

# Metadata
meta <- read_tsv(file = "~/SFA/data_amplicon/SFA2_full/SFA2_full_metadata.tsv") %>% 
  select(Sample, Type, Innoculant, DOC, Replicate, Day)

# Merge and average technical replicates
doc_meta <- inner_join(doc, meta) %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(mean_doc = mean(mean)) %>% 
  ungroup() %>% 
  mutate(Innoculant = as.character(Innoculant))

# Join with growth estimates
growth_summary <- growth %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(mean_k = mean(k),
            mean_start_day = mean(start_day),
            mean_end_day = mean(end_day),
            total_start_abund = sum(start_abund),
            total_end_abund = sum(end_abund),
            total_change_abund = sum(change_abund)) %>% 
  ungroup()

doc_growth <- inner_join(doc_meta, growth_summary)
```

```{r}
# k
doc_growth %>% 
  ggplot(aes(x=mean_doc, y=mean_k, color=DOC)) +
  geom_point() +
  theme_test()

# start day
doc_growth %>% 
  ggplot(aes(x=mean_doc, y=mean_start_day, color=DOC)) +
  geom_point() +
  theme_test()

# start abundance
doc_growth %>% 
  ggplot(aes(x=mean_doc, y=total_start_abund, color=DOC)) +
  geom_point() +
  theme_test()

# end abundance
doc_growth %>% 
  ggplot(aes(x=mean_doc, y=total_end_abund, color=DOC)) +
  geom_point() +
  theme_test()

# change abundance
doc_growth %>% 
  ggplot(aes(x=mean_doc, y=total_change_abund, color=DOC)) +
  geom_point() +
  theme_test()
```

Not seeing anything worth persuing further.

# ARCHIVE

ASV-level pretty 
* sure this is a bad idea!!! incorporates pseudoreps
* distributions are the way to go I think
* need to think about how distributions handle replication
* or maybe mixed linear effects model could handle this? make innoculant a random effect? inforunately ASV should and can't be made random effect

```{r, eval=FALSE}
# Growth rate
growth %>% 
  ggplot(aes(x=DOC, y=k)) +
  geom_boxplot() +
  geom_jitter() +
  theme_test()

# Start day
growth %>% 
  ggplot(aes(x=DOC, y=start_day)) +
  geom_boxplot() +
  geom_jitter() +
  theme_test()

# Change in abundance
growth %>% 
  ggplot(aes(x=DOC, y=log(change_abund))) +
  geom_boxplot() +
  geom_jitter() +
  theme_test()

# Start abundance
growth %>% 
  ggplot(aes(x=DOC, y=log(start_abund))) +
  geom_boxplot() +
  geom_jitter() +
  theme_test()

# End abundance
growth %>% 
  ggplot(aes(x=DOC, y=log(end_abund))) +
  geom_boxplot() +
  geom_jitter() +
  theme_test()
```

