---
title: "WIP_death_explore"
author: "Cassandra Wattenburger"
date: "11/18/2021"
output: html_document
---


```{r}
# Clear working directory, load in packages, generate package info
rm(list=ls())

library("tidyverse")
library("trelliscopejs")

sessionInfo()
```

# Import data

Files produced in 03_SFA2_deathestimates.Rmd
* death estimates
* normalized abundance data for time series with estimated death

```{r}
# death
death <- readRDS("../data_intermediate/SFA2_death_estimates.rds")

# Taxonomic information
tax <- readRDS("../data_intermediate/SFA2_tax_normalized.rds")
```

# Exploration

### Numbers of dying bacteria

```{r}
# Number of taxa dying
num_grow <- death %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(num_grew = n()) %>% 
  ungroup()

num_grow

num_grow %>% 
  ggplot(aes(x=DOC, y=num_grew)) +
  geom_boxplot() +
  labs(x="DOC \"phenotype\"", y="Number of death estimated taxa") +
  theme_test() +
  theme()
```

### Phylogenetic membership:

```{r}
# Merge death estimates with taxonomy
death_tax <- inner_join(death, tax) 

# Graph phylogenetic membership
death_tax %>% 
  group_by(Phylum, DOC, Innoculant) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  arrange(count) %>% 
  ggplot(aes(x=count, y=reorder(Phylum, -count), color=DOC)) +
  geom_boxplot() +
  #facet_wrap(~DOC) +
  theme_test() +
  theme(axis.text.x = )
```

### Death parameters:

* specific death rate (k)
* start day
* change in abundance
* start abundance
* end abundance

### Distributions of parameters

Specific death rates

```{r}
# Distribution of death rates
death %>%
  ggplot(aes(x=slope, color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="Specific death rate (slope)", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

death %>%
  ggplot(aes(x=slope, color=Innoculant)) +
  geom_density(alpha=0.5) +
  labs(x="Specific death rate (k)", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

Change in relational abundance

```{r}
# Distribution of change in abundance during death
death %>%
  ggplot(aes(x=log(change_abund), color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Change in relational abundance during death", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

death %>%
  ggplot(aes(x=log(change_abund), color=Innoculant, linetype=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Change in relational abundance during death", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

Start relational abundance

```{r}
# Distribution of change in abundance during death
death %>%
  ggplot(aes(x=log(start_abund), color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Starting relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

death %>%
  ggplot(aes(x=log(start_abund), color=Innoculant)) +
  geom_density(alpha=0.5) +
  labs(x="ln Starting in relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

End relational abundance

```{r}
# Distribution of change in abundance during death
death %>%
  ggplot(aes(x=log(end_abund), color=DOC)) +
  geom_density(alpha=0.5) +
  labs(x="ln Ending relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

death %>%
  ggplot(aes(x=log(end_abund), color=Innoculant)) +
  geom_density(alpha=0.5) +
  labs(x="ln Ending relational abundance", y="Kernel Density") +
  theme_test() +
  theme(#legend.position="none",
        title=element_text(size=12),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))
```

### Community-level

```{r}
# Summarize parameters
death_summary <- death %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(mean_slope = mean(slope),
            mean_start_day = mean(start_day),
            mean_end_day = mean(end_day),
            mean_end_abund = mean(end_abund),
            total_end_abund = sum(end_abund),
            mean_change_abund = mean(change_abund),
            total_change_abund = sum(change_abund)) %>% 
  ungroup()
```

```{r}
# Mean slope
death_summary %>% 
  ggplot(aes(x=DOC, y=mean_slope)) +
  geom_point() +
  theme_test()

# Mean start day
death_summary %>% 
  ggplot(aes(x=DOC, y=mean_start_day)) +
  geom_point() +
  theme_test()

# Mean change abundance
death_summary %>% 
  ggplot(aes(x=DOC, y=mean_change_abund)) +
  geom_point() +
  theme_test()

# Sum of change abundance
death_summary %>%
  ggplot(aes(x=DOC, y=total_change_abund)) +
  geom_point() +
  theme_test()

# Mean end abund
death_summary %>% 
  ggplot(aes(x=DOC, y=mean_end_abund)) +
  geom_point() +
  theme_test()

# Sum of end abundance
death_summary %>% 
  ggplot(aes(x=DOC, y=total_end_abund)) +
  geom_point() +
  theme_test()
```

# Relationship to DOC measurement

Import DOC measurements

```{r}
# DOC measurements
doc <- readRDS("../data_DOC/SFA2/SFA2_doc.rds") %>% 
  select(everything(), Sample=sample_id) %>% 
  mutate(Sample = as.numeric(Sample))

# Metadata
meta <- read_tsv(file = "../data_amplicon/SFA2_full/SFA2_full_metadata.tsv") %>% 
  select(Sample, Type, Innoculant, DOC, Replicate, Day)

# Merge and average technical replicates
doc_meta <- inner_join(doc, meta) %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(doc_avg = mean(mean)) %>% 
  ungroup() %>% 
  mutate(Innoculant = as.character(Innoculant))

# Join with death estimates
death_summary <- death %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(mean_slope = mean(slope),
            mean_start_day = mean(start_day),
            mean_end_day = mean(end_day),
            total_start_abund = sum(start_abund),
            total_end_abund = sum(end_abund),
            total_change_abund = sum(change_abund)) %>% 
  ungroup()

doc_death <- inner_join(doc_meta, death_summary)
```

```{r}
# k
doc_death %>% 
  ggplot(aes(x=doc_avg, y=mean_slope, color=DOC)) +
  geom_point() +
  theme_test()

# start day
doc_death %>% 
  ggplot(aes(x=doc_avg, y=mean_start_day, color=DOC)) +
  geom_point() +
  theme_test()

# start abundance
doc_death %>% 
  ggplot(aes(x=doc_avg, y=total_start_abund, color=DOC)) +
  geom_point() +
  theme_test()

# end abundance
doc_death %>% 
  ggplot(aes(x=doc_avg, y=total_end_abund, color=DOC)) +
  geom_point() +
  theme_test()

# change abundance
doc_death %>% 
  ggplot(aes(x=doc_avg, y=total_change_abund, color=DOC)) +
  geom_point() +
  theme_test()
```