---
title: "Untitled"
author: "Cassandra Wattenburger"
date: "1/20/2022"
output: html_document
---

```{r}
# Clear working directory, load in packages, generate package info
rm(list=ls())

library("tidyverse")

sessionInfo()
```

# Import data

Files produced in 02_SFA2_growthestimates.Rmd, 03_SFA2_deathestimates.Rmd.

```{r}
# growth
growth <- readRDS("../data_intermediate/SFA2_growth_estimates.rds")

# death
death <- readRDS("../data_intermediate/SFA2_death_estimates.rds")

# Taxonomic information
tax <- readRDS("../data_intermediate/SFA2_tax_normalized.rds")
```

# Overlapping ASVs - growth

```{r}
asv_overlap <- growth %>% 
  select(DOC, ASV) %>% 
  unique() %>% 
  group_by(ASV) %>% 
  summarize(overlap = n()) %>% # 1 means not in both trts, two means in both trts
  filter(overlap == 2) %>% 
  select(ASV)

dim(asv_overlap)
```

There are 25 shared ASVs between high and low DOC phenotype communities. Who are they?

```{r}
asv_overlap <- inner_join(asv_overlap, tax)

asv_overlap %>% 
  arrange(Phylum)

asv_overlap %>% 
  group_by(Phylum) %>% 
  summarize(count = n()) %>% 
  arrange(-count)
```

Overlapping taxa mainly consist of Actinobacteria, Firmicutes, and Proteobacteria.

Change in normalized abundance was significantly different between treatments. Do overlapping taxa show same trend?

```{r}
growth %>% semi_join(asv_overlap, growth, by="ASV") %>% 
  group_by(DOC, Innoculant) %>% 
  summarize(mean_chgabund = mean(end_abund - start_abund)) %>% 
  ungroup() %>% 
  ggplot(aes(x=DOC, y=mean_chgabund)) + 
  geom_point() + 
  theme_test()
```

Yes, these seem to recreate the difference between treatments.



