---
title: "Untitled"
author: "Cassandra Wattenburger"
date: "4/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
# Clear working directory, load in packages, generate package info
rm(list=ls())

library("tidyverse")

sessionInfo()
```

# Notes

# Import data

* growth and death estimates w/PAPRICA predictions

```{r}
# Growth
growth <- readRDS("../data_intermediate/SFA2_growth_estimates_pap.rds")

# Death
death <- readRDS("../data_intermediate/SFA2_death_estimates_pap.rds")
```

# Growth vs growth

k vs lag time:

```{r}
growth %>% 
  ggplot(aes(x=start_day, y=k)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Inoculant) +
  theme_test() 
```

k vs change in abundance:

```{r}
growth %>% 
  ggplot(aes(x=log(change_abund), y=k)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Inoculant) +
  theme_test() 
```

k vs length:

```{r}
growth %>% 
  mutate(length=end_day - start_day) %>% 
  ggplot(aes(x=length, y=k)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Inoculant) +
  theme_test() 
```

Length vs change in abundance:

```{r}
growth %>% 
  mutate(length=end_day - start_day) %>% 
  ggplot(aes(x=length, y=log(change_abund))) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~Inoculant) +
  theme_test() 
```

Does combo of positive relationship between k and change in abundance with negative relationship between k and length of growth indicate increasing efficiency with increasing growth rate?

Faster growth rate <-> shorter period of growth & greater change in abundance

Would need CO2 respiration and C consumption data for individual taxa to further validate.

