---
title: "Untitled"
author: "Cassandra Wattenburger"
date: "3/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
# Clear working directory, load in packages, generate package info
rm(list=ls())

library("tidyverse")

sessionInfo()
```

# Notes


* Growth parameters
  * specific growth rate (k)
  * start day
  * end day
  * length of growth
  * start abundance
  * end abundance
  * change in abundance
* Other
  * 16S copy number
  * genome size
  * phi (genome plasticity)
  * number of pathways
  * number of catalogued enzymes
  
* abundance data has already been corrected for via n16S prediction

# Import data

* growth estimates w/paprica predictions
* normalized abundance data for time series with estimated growth

```{r}
# Growth
growth <- readRDS("../data_intermediate/SFA2_growth_estimates_pap.rds")

# Normalized abundance data
norm <- readRDS("../data_intermediate/SFA2_norm_grestimated.rds")

# Taxonomic information
tax <- readRDS("../data_intermediate/SFA2_taxonomy.rds")
```

# Phylogenetic membership

Overall:

```{r}
# Merge growth estimates with taxonomy
growth_tax <- inner_join(growth, tax) 

# Graph phylogenetic membership
growth_tax %>% 
  group_by(Phylum, Inoculant) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  group_by(Phylum) %>% 
  summarize(count_mean = mean(count), count_sd = sd(count)) %>% 
  ungroup() %>% 
  arrange(count_mean)
```

# Inoculant overviews

Plots:

```{r}
# specific growth rate (k)
growth %>% 
  ggplot(aes(x=Inoculant, y=k)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# start abundance
growth %>% 
  ggplot(aes(x=Inoculant, y=log(start_abund))) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test() 

# end abundance
growth %>% 
  ggplot(aes(x=Inoculant, y=log(end_abund))) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# change abundance
growth %>% 
  ggplot(aes(x=Inoculant, y=log(change_abund))) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# lag
growth %>% 
  ggplot(aes(x=Inoculant, y=start_day)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# end day
growth %>% 
  ggplot(aes(x=Inoculant, y=end_day)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# length
growth %>% 
  mutate(length = end_day - start_day) %>% 
  ggplot(aes(x=Inoculant, y=length)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# n16S
growth %>% 
  ggplot(aes(x=Inoculant, y=n16S)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# phi
growth %>% 
  ggplot(aes(x=Inoculant, y=phi)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# pathways
growth %>% 
  ggplot(aes(x=Inoculant, y=npaths_actual)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()

# enzymes
growth %>% 
  ggplot(aes(x=Inoculant, y=nec_actual)) +
  geom_boxplot() +
  geom_jitter(alpha=0.5) +
  theme_test()
```

For start day there is a batch that starts growing on day 0, a prominent gap, and then another batch that starts growing at varying points afterwards. Is this a wave? It seems to correspond to the shift towards homogenizing selection?

"succession is a result of waves of births and deaths"

what is the difference between assembly and succession?






